Convergence rate of MAP estimates \\ for the exponential family

### 2021-03-18 main.tex
Lots of modifications - add bias variance section

@@

\begin{example}
	These frames contain the trailing example of this paper: a centered gaussian with unknown variance $\cN(0,\sigma^2)$.
	The density of a centered normal variable is
\begin{align}
	p(x) = \inv{\sqrt{2\pi \sigma^2}} e^{-\frac{x^2}{2 \sigma^2}} \; .
\end{align}
Defining $T(X)=X^2$ as the sufficient statistic, we get natural parameter $\natp = -\inv{2 \sigma^2} <0$, and mean parameter $\mu=\E[T(X)] = \sigma^2 >0$. 
Mean and natural parameters are roughly inverse of each other
\begin{align}
	\natp = -\inv{2 \mu}\; .
\end{align}
Now we can match the log-likelihood with the exponential family template to get the log-partition function.
\begin{align}
	\log p(x) = \frac{x^2}{2 \sigma^2} - \half \log(2 \pi \sigma^2 ) 
	= x^2 \natp - \logpart(\natp) \\
	\implies \logpart (\natp) = \half \left( -\log(-\natp)  + \log(\pi) \right)
\end{align}
\end{example}

\paragraph{Duality}
The logpartition function $\logpart$ verifies the two following identities
\begin{align}
    \nabla\logpart(\natp) &=  \expect[p(X|\natp)]{T(X)} =: \meanp \\
    \nabla^2 \logpart(\natp) &= \Cov_\natp[T(X)] > 0
\end{align}
where $\meanp$ is called the mean parameter.
If the sufficient statistic $T$ is minimal, then the log-partition function $\logpart$ is strictly convex and its gradient $\nabla \logpart$ is a bijection between natural parameters $\natp$ and mean parameters $\mu$.
The second identity entails that $\logpart$ is strictly-convex. 
At this point it is useful to introduce the \href{https://en.wikipedia.org/wiki/Convex_conjugate}{convex conjugate} (aka Fenchel-Legendre transform) of the logpartition function
\begin{align}
	\conj(\mu) = \langle \mu, \natp \rangle - \logpart(\natp) \; .
\end{align}
It turns out that $\conj$ matches the common notion of \textit{entropy} in information theory, so we will call it entropy.
If $\logpart$ is strictly convex, then its gradient is strictly monotone, so it is a bijection, and its inverse is the gradient of its dual $\nabla\conj \circ \nabla\logpart(\natp) = \natp$ (cf Fig.~\ref{fig:duality}).
For a full review of exponential families and their duality, see \citet[Chapter 3]{wainwright2008graphical}.
\begin{figure}[h]
	\centering
	\includegraphics{figs/duality}
	\caption{The gradient of the log-partition function and its dual, $(\nabla \logpart, \nabla \conj)$, form a bijection between the natural and mean parameters $\natp, \meanp$. Figure reproduced from \citet{kunstner2020homeomorphic}.}
	\label{fig:duality}
\end{figure}
\begin{example}
		For $\cN(0,\sigma^2)$,we can use the formula $\conj(\mu) = \mu \natp - \logpart(\natp)$ to get the entropy
\begin{align}
	\conj(\mu) = \half\left( -\log(\mu) + \log\frac{\pi}{2} - 1 \right) \; .
\end{align}
\end{example}

@@

\section{Problem}
For a well-specified model, the suboptimality on the population log-likelihood is exactly the KL between our current model and the true distribution
\begin{align}
    \expect[X\sim p(.|\natp^*)]{-\log p(X|\natp) + \log p(X|\natp^*) }
	= \KL( p(.|\natp^*) ; p(.|\natp)) \; .
\end{align}
For the exponential family, the KL is also the Bregman divergence induced by the log-partition function (with switched arguments)
\begin{align}
	\KL( p(.|\natp^*) ; p(.|\natp)) 
	= \bregman (\natp ; \natp^*)  \; .
\end{align}
There is a general relationship between Bregman divergences and convex conjugates (notice the argument switching)
\begin{align}
	\bregman (\natp ; \natp^*)
    = \logpart(\natp) - \langle \natp , \mu^* \rangle + \conj(\mu^*)
    = \bregmanconj ( \meanp^* ; \meanp)
\end{align}
so in the end the suboptimality is a divergence, which can either be seen as a KL between distributions, as a divergence between natural parameters, or as a divergence between mean parameters
\begin{align}
\boxed{
	\KL( p(.|\natp^*) ; p(.|\natp))
    = \bregman (\natp ; \natp^*)
    = \bregmanconj ( \meanp^* ; \meanp) \; .
}
\end{align}
\begin{important}
The question is: how does this quantity behave when $\natp$ is the maximum-likelihood or the MAP estimate ? Can we get bounds -- in expectation or high-probability -- on the following quantities ?
\begin{align}
	\label{eq:bregmanMLE}
	\bregmanconj \left (\E [T(X)] ;  \inv{n}  \smallsum_i T(X_i) \right ) \; , \\
	\label{eq:bregmanMAP}
	\bregmanconj \left (\E [T(X)] ; \frac{n_0 \mu_0 + \smallsum_i T(X_i)}{n_0+n} \right ) \; ,
\end{align}
where the randomness is coming from the dataset $X_1, \dots, X_n$.
\end{important}

@@

\begin{example}
	\paragraph{Exact computation}
	The MLE of $\cN(0,\mu_*)$ is $\mu_n = \sum_i X_i^2 /n$. The ratio to the optimum  $\frac{\mu_n}{\mu_*}$follows a \href{https://en.wikipedia.org/wiki/Chi-square_distribution}{Chi-square distribution} with $n$ degrees of freedom $\chi^2(n)$ divided by $n$. Its inverse $\frac{\mu_*}{ \mu_n} $ follows an \href{https://en.wikipedia.org/wiki/Inverse-chi-squared_distribution}{inverse Chi-square distribution} with expectation 
	\begin{align}
		\expect{\frac{\mu_*}{ \mu_n} } =
		\begin{cases}
			\frac{n}{n-2} \ \text{ if } n>2, \\
			+\infty \  \text{ otherwise. }
		\end{cases}
	\end{align}
	There is also a closed form solution for the expected logarithm of a Chi-squared\citep{pav2015moments}
	\begin{align}
		\expect{\log \frac{\mu_n}{\mu_*}} = \log(2) + \psi(\half[n]) - \log(n)
	\end{align}
	where $\psi$ is the \href{https://en.wikipedia.org/wiki/Digamma_function}{digamma function.}
	Consequently the suboptimality of the MLE has a closed form solution
	\begin{align}
	\expect{\bregmanconj( \mu_*; \mu_n) }
	&= \half \expect{\frac{\mu_*}{ \mu_n} - 1 + \log \left(\frac{\mu_n}{\mu_*} \right) } \\
	& =
	\begin{cases}
		\half \left ( \frac{2}{n-2} +\psi(\half[n]) - \log(\half[n]) \right )  \ \text{ if } n>2, \\
			+\infty \  \text{ otherwise. }
	\end{cases}
	\end{align}
	It is surprising that we need more than $3$ samples for the loss to have a bounded expectation. 
	It is also very pleasant that the optimal value $\mu_*$ does not appear in this rate, so that the convergence rate is independent of the actual solution.
	When the expectation is finite, we can make its formula intelligible thanks to \href{https://en.wikipedia.org/wiki/Digamma_function#Inequalities}{bounds on the digamma function}
	\begin{align}
		-\inv{x} \leq \psi(x) - \log(x) \leq -\inv{2x}	\; .
	\end{align}
	  For $n\geq 3$,
	\begin{align}
		\inv{n-2} - \inv{n}
		&\leq \expect{\bregmanconj( \mu_*; \mu_n) }
		\leq \inv{n-2} - \inv{2 n} \\
		\iff
			\frac{2}{n(n-2)}
			&\leq \expect{\bregmanconj( \mu_*; \mu_n) }
			\leq \inv{2n} +\frac{2}{n(n-2)}
			\label{eq:tight_bounds}
	\end{align}
	so we get a $O(n^{-2})$ lower bound and a $O(n^{-1}) + O(n^{-2})$ upper bound.
	Using the \href{https://en.wikipedia.org/wiki/Digamma_function#Recurrence_formula_and_characterization}{Stirling series}, we can also get an asymptotic approximation
	\begin{align}
		\expect{\bregmanconj( \mu_*; \mu_n) }
		\in \inv{2n} + \frac{11}{6 n^2} + \frac{4}{n^2(n-2)} + O(n^{-4}) \; .
	\end{align}
	Unfortunately, formulas for the MAP seem to be much less nice. There is also a closed form for the multivariate generalization, thanks to the \href{https://en.wikipedia.org/wiki/Inverse-Wishart_distribution}{inverse Wishart distribution} and the \href{https://en.wikipedia.org/wiki/Wishart_distribution#Log-expectation}{expectation of the log-determinant of a Wishart}. Again the expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension. We do not perform these calculations for now as we want to focus on the simplest possible examples.
\end{example}

@@

@@

We know that asymptotically, the convergence rate is dominated by $O(n^{-1})$ variance term. To this we may add a bias term in $O(n^{-2})$. 
We also know from the $\cN(0,\mu_*)$ example that the MLE can have infinite loss when there is not enough samples. This is not the case of the MAP which always has a bounded loss, since it can never reach the boundary of the marginal polytope (the domain of $\mu$). Thus we may have a bias term which is infinite for $n_0=0$ (eg no prior, MLE), and that decreases as $n_0$ increases, except that the convergence speed gets worse and worse for larger $n_0$, eg
\begin{align}
    \frac{d}{2n} +  \log \left (\inv{n_0} \right) \left(1+ \frac{n}{n_0} \right)^{-2} \bregman(\natp_*; \natp_0)
\end{align}
We also know that this expression should dominate the asymptote.
Only the variance is responsible for the infinite behavior, except it gets controlled when $n_0>0$.

I really want to get a bound for the MAP of a Gaussian variance, to get inspiration from its shape.

@@

\section{Gaussian Variance}
\subsection{A big calculus in this specific setting}

@@

>>>>

%% Faster processing
\renewcommand\appendix{\end{document}}

@@

\begin{example}
	These frames contain the trailing example of this paper: a centered gaussian with unknown variance $\cN(0,\sigma^2)$.
	The density of a centered normal variable is
\begin{align}
	p(x) = \inv{\sqrt{2\pi \sigma^2}} e^{-\frac{x^2}{2 \sigma^2}} \; .
\end{align}
Defining $T(X)=X^2$ as the sufficient statistic, we get natural parameter $\natp = -\inv{2 \sigma^2} <0$, and mean parameter $\mu=\E[T(X)] = \sigma^2 >0$. 
Mean and natural parameters are roughly inverse of each other
\begin{align}
	\natp = -\inv{2 \mu}\; .
\end{align}
Now we can match the log-likelihood with the exponential family template to get the log-partition function.
\begin{align}
	\log p(x) = - \frac{x^2}{2 \sigma^2} - \half \log(2 \pi \sigma^2 ) 
	= x^2 \natp - \logpart(\natp) \\
	\implies \logpart (\natp) = - \half \log(-\natp)  + \half \log(\pi) 
\end{align}
\end{example}

\paragraph{Duality}
The logpartition function $\logpart$ verifies the two following identities
\begin{align}
    \nabla\logpart(\natp) &=  \expect[p(X|\natp)]{T(X)} =: \meanp \\
    \nabla^2 \logpart(\natp) &= \Cov_\natp[T(X)] > 0
\end{align}
where $\meanp$ is called the mean parameter.
If the sufficient statistic $T$ is minimal, then the log-partition function $\logpart$ is strictly convex and its gradient $\nabla \logpart$ is a bijection between natural parameters $\natp$ and mean parameters $\mu$.
The second identity entails that $\logpart$ is strictly-convex. 
At this point it is useful to introduce the \href{https://en.wikipedia.org/wiki/Convex_conjugate}{convex conjugate} (aka Fenchel-Legendre transform) of the logpartition function
\begin{align}
	\conj(\mu) = \langle \mu, \natp \rangle - \logpart(\natp) \; .
\end{align}
It turns out that $\conj$ matches the common notion of \textit{entropy} in information theory, so we will call it entropy.
If $\logpart$ is strictly convex, then its gradient is strictly monotone, so it is a bijection, and its inverse is the gradient of its dual $\nabla\conj \circ \nabla\logpart(\natp) = \natp$ (cf Fig.~\ref{fig:duality}).
For a full review of exponential families and their duality, see \citet[Chapter 3]{wainwright2008graphical}.
\begin{figure}[h]
	\centering
	\includegraphics{figs/duality}
	\caption{The gradient of the log-partition function and its dual, $(\nabla \logpart, \nabla \conj)$, form a bijection between the natural and mean parameters $\natp, \meanp$. Figure reproduced from \citet{kunstner2020homeomorphic}.}
	\label{fig:duality}
\end{figure}
\begin{example}
For $\cN(0,\sigma^2)$,we can use the formula $\conj(\mu) = \mu \natp - \logpart(\natp)$ to get the entropy
\begin{align}
	\conj(\mu) = \half\left( -\log(\mu) + \log\frac{\pi}{2} - 1 \right) \; .
\end{align}
We can also take gradient and derivative of $\logpart(\natp)$ to retrieve the mean and covariance of the sufficient statistic $X^2$
\begin{align}
	\nabla\logpart(\natp) &= \frac{-1}{2 \natp} = \sigma^2 = \mu = \E[X^2] \\
	\nabla^2\logpart(\natp) &= \frac{1}{2 \natp^2} = 2 \sigma^4 = 2 \mu^2 = \Var(X^2) 
\end{align}
which we confirm thank to wikipedia since $\E[X^4] = 3 \sigma^4$ and thus $\Var(X^2) = \E[X^4] - \E[X^2]^2 = 3 \sigma^4 - \sigma^4 = 2 \sigma^4$.
\end{example}

@@

\section{Problem}
For a well-specified model, the suboptimality on the population log-likelihood is exactly the KL between our current model and the true distribution
\begin{align}
    \expect[X\sim p(.|\natp^*)]{-\log p(X|\natp) + \log p(X|\natp^*) }
	= \KL( p(.|\natp^*) ; p(.|\natp)) \; .
\end{align}
For the exponential family, the KL is also the Bregman divergence induced by the log-partition function (with switched arguments)
\begin{align}
	\KL( p(.|\natp^*) ; p(.|\natp)) 
	= \bregman (\natp ; \natp^*)  \; .
\end{align}
There is a general relationship between Bregman divergences and convex conjugates (notice the argument switching)
\begin{align}
	\bregman (\natp ; \natp^*)
    = \logpart(\natp) - \langle \natp , \mu^* \rangle + \conj(\mu^*)
    = \bregmanconj ( \meanp^* ; \meanp)
\end{align}
so in the end the suboptimality is a divergence, which can either be seen as a KL between distributions, as a divergence between natural parameters, or as a divergence between mean parameters
\begin{align}
\boxed{
	\KL( p(.|\natp^*) ; p(.|\natp))
    = \bregman (\natp ; \natp^*)
    = \bregmanconj ( \meanp^* ; \meanp) \; .
}
\end{align}
\begin{important}
The question is: how does this quantity behave when $\natp$ is the maximum-likelihood or the MAP estimate ? Can we get bounds on the following quantities
\begin{align}
	\label{eq:bregmanMLE}
	\expect[X_i\sim \natp^*]{\bregmanconj \left (\E [T(X)] ;  \inv{n}  \smallsum_i T(X_i) \right )} \leq \ ? \; , \\
	\label{eq:bregmanMAP}
	\expect[X_i\sim \natp^*]{\bregmanconj \left (\E [T(X)] ; \frac{n_0 \mu_0 + \smallsum_i T(X_i)}{n_0+n} \right )} \leq \ ? \; ,
\end{align}
where the outer expectation is on the dataset $X_1, \dots, X_n$?
\end{important}

@@

\begin{example}
	\paragraph{Exact computation}
	The MLE of $\cN(0,\mu_*)$ is $\mu_n = \sum_i X_i^2 /n$. The ratio to the optimum  $\frac{\mu_n}{\mu_*}$follows a \href{https://en.wikipedia.org/wiki/Chi-square_distribution}{Chi-square distribution} with $n$ degrees of freedom $\chi^2(n)$ divided by $n$. Its inverse $\frac{\mu_*}{ \mu_n} $ follows an \href{https://en.wikipedia.org/wiki/Inverse-chi-squared_distribution}{inverse Chi-square distribution} with expectation 
	\begin{align}
		\expect{\frac{\mu_*}{ \mu_n}  - 1} 
		= \expect{\frac{n}{\chi^2(n)} - 1} =
		\begin{cases}
			\frac{n}{n-2} -1 = \frac{2}{n-2} \ \text{ if } n>2, \\
			+\infty \  \text{ otherwise. }
		\end{cases}
	\end{align}
	There is also a closed form solution for the expected logarithm of a Chi-squared\citep{pav2015moments}
	\begin{align}
		\expect{\log \frac{\mu_n}{\mu_*}} = \psi(\half[n]) - \log(\half[n])
	\end{align}
	where $\psi$ is the \href{https://en.wikipedia.org/wiki/Digamma_function}{digamma function.}
	Consequently the suboptimality of the MLE has a closed form solution
	\begin{align}
	\expect{\bregmanconj( \mu_*; \mu_n) }
	&= \half \expect{\frac{\mu_*}{ \mu_n} - 1 + \log \left(\frac{\mu_n}{\mu_*} \right) } \\
	& =
	\begin{cases}
		\half \left ( \frac{2}{n-2} +\psi(\half[n]) - \log(\half[n]) \right )  \ \text{ if } n>2, \\
			+\infty \  \text{ otherwise. }
	\end{cases}
	\end{align}
	It is surprising that we need more than $3$ samples for the loss to have a bounded expectation. 
	It is also very pleasant that the optimal value $\mu_*$ does not appear in this rate, so that the convergence rate is independent of the actual solution.
	When the expectation is finite, we can make its formula intelligible thanks to \href{https://en.wikipedia.org/wiki/Digamma_function#Inequalities}{bounds on the digamma function}
	\begin{align}
		-\inv{x} \leq \psi(x) - \log(x) \leq -\inv{2x}	\; .
	\end{align}
	  For $n\geq 3$,
	\begin{align}
		\inv{n-2} - \inv{n}
		&\leq \expect{\bregmanconj( \mu_*; \mu_n) }
		\leq \inv{n-2} - \inv{2 n} \\
		\iff
			\frac{2}{n(n-2)}
			&\leq \expect{\bregmanconj( \mu_*; \mu_n) }
			\leq \inv{2n} +\frac{2}{n(n-2)}
			\label{eq:tight_bounds}
	\end{align}
	so we get a $O(n^{-2})$ lower bound and a $O(n^{-1}) + O(n^{-2})$ upper bound.
	Using the \href{https://en.wikipedia.org/wiki/Digamma_function#Recurrence_formula_and_characterization}{Stirling series}, we can also get an asymptotic approximation
	\begin{align}
		\expect{\bregmanconj( \mu_*; \mu_n) }
		\in \inv{2n} + \frac{11}{6 n^2} + \frac{4}{n^2(n-2)} + O(n^{-4}) \; .
	\end{align}
	Unfortunately, formulas for the MAP seem to be much less nice. There is also a closed form for the multivariate generalization, thanks to the \href{https://en.wikipedia.org/wiki/Inverse-Wishart_distribution}{inverse Wishart distribution} and the \href{https://en.wikipedia.org/wiki/Wishart_distribution#Log-expectation}{expectation of the log-determinant of a Wishart}. Again the expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension. We do not perform these calculations for now as we want to focus on the simplest possible examples.
\end{example}

@@

\section{Bias-Variance Decomposition}
Using Jensen inequality, we have the following analog of the bias-variance decomposition:

\newcommand{\bias}[1]{\textcolor{Red}{\underbrace{\textcolor{black}{#1}}_{\text{bias}}}}
\newcommand{\variance}[1]{\textcolor{Green}{\underbrace{\textcolor{black}{#1}}_{\text{variance}}}}

\begin{align}
	& \expect{\bregmanconj(\mu^* ; \hat \mu_n)} \\
	& \leq \bias{\bregmanconj(\mu^*; \mu_n)}
	+ \frac{n_0}{n+n_0} \langle  \bias{\mu_* - \mu_0} ; \variance{\natp_n - \E[\hat \natp_n]} \rangle 
	+ \frac{1}{n+n_0} \smallsum_i \variance{\expect{\langle \hat \natp_n ; T(X_i) - \mu^* \rangle } }
\end{align}
where $\hat \mu_n$ is the MAP estimate,
$\mu_n = \expect{\hat \mu_n} = \frac{n_0 \mu_0 + n \mu^*}{n_0 + n}$,
and $\theta_n = \nabla \conj(\mu_n)$.

@@

We know that asymptotically, the convergence rate is dominated by $O(n^{-1})$ variance term. To this we may add a bias term in $O(n^{-2})$. 
We also know from the $\cN(0,\mu_*)$ example that the MLE can have infinite loss when there is not enough samples. This is not the case of the MAP which always has a bounded loss, since it can never reach the boundary of the marginal polytope (the domain of $\mu$). 
I really want to get a bound for the MAP of a Gaussian variance, to get inspiration from its shape. 
Thus we may have a bias term which is infinite for $n_0=0$ (eg no prior, MLE), and that decreases as $n_0$ increases, except that the convergence speed gets worse and worse for larger $n_0$, eg
\begin{align}
    \frac{d}{2n} +  \log \left (\inv{n_0} \right) \left(1+ \frac{n}{n_0} \right)^{-2} \bregman(\natp_*; \natp_0)
\end{align}
Besides this bias term should be visible from a simple mirror descent, with no noise or variance. Yet it is not trivial to get a $O(n^{-2})$ rate in this setting. See note by Frederik.

@@

\subsection{Convergence of Mirror Descent}
To get the descent lemma, apply successively relative smoothness, 3 points lemma, and relative strong convexity on $f$. Note that the strong convexity constant $m$ may be zero. Recall that $\ell_f(\natp, \natp^{t}) = f(\natp^t) + \langle \nabla f (\natp^t) ; \natp - \natp^t \rangle $ is the linearization of $f$ at $\natp^t$ evaluated in $\natp$. 
\begin{align}
	f(\natp^{t+1}) 
	&\leq \ell_f(\natp^{t+1}, \natp^{t})  + L \bregman(\natp^{t+1}; \natp^{t})
	& \text{(Relative $L$-smoothness)} \\
	& \leq \ell_f(\natp^{*}, \natp^{t})  + L \bregman(\natp^{*}; \natp^{t}) -L \bregman(\natp^{*}; \natp^{t+1}) 
	& \text{(3 Points Lemma w. $x=\natp^*$)} \\
	& \leq f(\natp^*) + (L - m) \bregman(\natp^{*}; \natp^{t}) -L \bregman(\natp^{*}; \natp^{t+1}) 
	& \text{(Relative $m$-strong convexity)}
\end{align}
Applying the 3 points lemma with \TODO{f}

@@

\section{Vivien's approach}
\subsection{A big calculus in this specific setting}
